Novel improvements in the Material Point Method to face dynamic problems: Local-maximum entropy approximation and an explicit predictor-corrector scheme. \thanks{Funding: This study was funded by Agustín de Betancourt Foundation (grant number 262390106114).}

### 2020-03-02 main.tex
cambios 1-02-2020

\title{Suitability of the local-maximum entropy approximation under
  the Material Point Method framework for dynamic problems. \thanks{Funding: This
    study was funded by Agustín de Betancourt Foundation (grant number
    262390106114).}
}
% \subtitle{}

\titlerunning{Suitability of the LME approximation under
  the MPM framework for dynamic problems.} % if too long for running head

@@

\section{Introduction}
\label{intro}
Since the proposal of the Material Point Method (\textbf{MPM}) by Sulsky {\it
  et al.} (1994)\cite{Sulsky1994}, as a generalization of
a particle-in-cell method \cite{Harlow1956} and a fluid implicit
particle (FLIP) method \cite{Brackbill1986}, MPM has been used to
simulate dynamic problems. However, in the simulations made with the
original MPM, there are numerical noises \cite{Tran2019e} when it is faced to solve
classic dynamic benchmarks like the one proposed by Dyka \& Ingel
(1995)\cite{Dyka1995}. These numerical inaccuracies give rise to the
development of other interpolation techniques to overcome this limitation. Some of this
alternatives techniques are the generalized interpolation material
point method (GIMP) Bardenhagen \& Kober (2004)\cite{Bardenhagen2004},
the dual domain material point method (DDMP) Zhang {\it et al.}
(2011)\cite{Zhang2011a}, the B-Spline MPM Tran {\it et al.}
(2019)\cite{Tran2019a}, the Conservative Taylor Least
Squares reconstruction Wobbes {\it et al.}
(2018)\cite{E_Wobbes_2018} and more recently the local
maximum-entropy (or local \textit{max-ent}) shape function first
introduced by Arroyo \& Ortiz (2006)\cite{Arroyo2006} has been tested
under the MPM framework by Wobbes {\it et al.}
(2020)\cite{Wobbes2020}. Alternatively to change the spatial
interpolation technique other researchers have overcome the well known
wiggle that tends to appear in this dynamic simulations by filtering
high-frequencies with an alternative time integration scheme, the
implicit GIMP (iGIMP) Charlton {\it et al.}
(2017)\nocite{Charlton2017} or the generalised-$\alpha$ scheme for the
MPM proposed by Tran \& Solowski (2019)\nocite{Tran2019e}.

The aim of this paper is to mitigate this spurious oscillations by the
employ of a suitable combination of the maximum-entropy (or local
\textit{max-ent}) shape functions, and the proposal of a explicit
predictor-corrector scheme. Local \textit{max-ent} shape functions was
first introduced by \cite{Arroyo2006}, it belongs to the class of convex 
approximation schemes and provides a seamless transition between
finite elements (\textbf{FE}) and mesh-free interpolations. The
approximation scheme is based on a compromise between minimizing the
width of the shape function support and maximizing the information
entropy of the approximation. The local \textit{max-ent} approximation
may be regarded as a regularization, or \textit{thermalization}, of
Delaunay triangulation which effectively resolves the degenerate cases
resulting from the lack of uniqueness of the triangulation. Local
\textit{max-ent} basis functions possess many desirable properties for
meshfree algorithms. First of all, they are entirely defined by the
nodal set and the domain of analysis. They are also non-negative,
satisfy the partition of unity property, and provide an exact
approximation for affine functions \cite{Arroyo2006}. This
approximation scheme has been proof to have a good performance under
the dynamic regime by other researchers like Navas {\it et al.}
(2018)\cite{Navas2018a} and Li {\it et al.} (2012)\cite{Li2012} for
Optimal Transportation Meshfree (\textbf{OTM}) method. On the other
hand, the time integration scheme here proposed is an accommodation of
the Newmark a-form with $\gamma = 0.5$ and $\beta = 0$ which is the central difference
explicit (CD). We have choose this method among other suitable
alternatives as Wilson {\it et al.} (1972)\nocite{Wilson1972} or Chung
\& Hulbert (1993)\nocite{Geranlized_alpha_1993} because it simplicity
and it good performance dealing with solids dynamics problems like those
presented in \cite{Navas2018a}.

The article is organized as follows. Section \ref{sec:Notation}
briefly reviews the notation here employed. Next, Section \ref{sec:derivation-mpm}
is devoted to present briefly the governing equations of the elastic
problem, the variational formulation and the Galerkin procedure. In Section \ref{sec:epc-algor-mpm} an explicit predictor-corrector
time integration scheme for the MPM is proposed. Section
\ref{sec:local-max-ent} briefly reviews the local \textit{max-ent}
basis functions here employed. In Section \ref{sec:dynamic-benchmark}
a comparison between both methods MPM and OTM employing \textit{max-ent} shape
function is performed as well a parametric study for the $\gamma$
parameter. Finally, conclusions and future research topics are exposed
in Section \ref{sec:conclusions}.

@@

\section{Derivation of the MPM}
\label{sec:derivation-mpm}
In this section, we provide an overview of the standard explicit MPM
algorithm as presented by \cite{Sulsky1994}. The method has three main
steps: a variational recovery process to project particle data to the
nodes of the grid, a Eulerian step where balance of momentum is
compute by solving internal and external forces in the nodes, and
finally the advection of the Lagrangian particles. We will first give
a introduction to the governing equations and then follow by an
algorithmic description of the process.

\subsection{Governing equations }
\label{sec:governing-equations}
In the MPM approach a continuum is considered. Let define a
$\varOmega$ domain occupied by an elastic body, and $\partial
\varOmega$ the boundary of the domain, where $\partial
\varOmega = \Gamma_d \bigcup \Gamma_n$ and $\Gamma_d \cap  \Gamma_n =
\emptyset$. Note that $\varGamma_d$ is reserved for Dirichlet boundary
conditions, and  $\varGamma_n$ for the Neumann. The behaviour of the
continuum can be described by a set of governing equations which are
(i) the balance of momentum equation
\begin{equation}
  \label{eq:Balance-momentum}
\rho \vect{a} = \Div{\tens{\sigma}} + \rho \vect{b},
\end{equation}
next (ii) the compatibility equation with ensure that the velocity
field is compatible with the Cauchy strain field,
\begin{equation}
  \label{eq:Compatibility-equation}
  \dot{\tens{\varepsilon}} = \GradS{\vect{v}} = \frac{1}{2}\left(
    \Grad{\vect{v}} +  \GradT{\vect{v}} \right),
\end{equation}
later (iii) the constitutive equation, which relates the rate of Cauchy strain
tensor with the rate of Cauchy stress tensor,
\begin{equation}
  \label{eq:Constitutive-equation}
\dot{\tens{\sigma}} = \tens{D} \colon \dot{\tens{\varepsilon}},
\end{equation}
finally we introduce (iv) the mass conservation, which if obtained by
setting to zero the total derivative of the density field,
\begin{equation}
  \label{eq:Rho-material-derivative}
  \frac{D \rho}{D t} = \dot{\rho} + \rho \Div\vect{v}} = 0.
\end{equation}

\subsection{Variational formulation}
\label{sec:variational-formulation}
To write the variational statement of the problem, let us define a
virtual velocity field such that
\begin{equation}
  \label{eq:Hilbert-space}
  \vect{v}^{\psi} \in \mathcal{H}^1_0(\Omega) = \{ \vect{v}^{\psi} \in
  \mathcal{H}^1 \mid \vect{v}^{\psi} = \vect{0}\ \text{on}\ \Gamma_d \}.
\end{equation}
And which satisfies that the Cauchy sequences are convergent in $\varOmega$
\begin{equation}
  \label{eq:cauchy-secuence}
  \Integral{3}{\vect{v}^{\psi}} < \infty\ \quad\text{and}\quad \Integral{3}{\Grad{\vect{v}^{\psi}}} < \infty
\end{equation}
The principle of virtual work states that the equilibrium solution to
the boundary value problem of elasticity is the function $\vect{v} \in
\mathcal{H}^1_0$ such that, for $\vect{v}^{\psi} \in
\mathcal{H}^1_0$, the following holds:
\begin{equation}
  \label{eq:BalanceMomentum_wf}
  \Integral{3}{\rho\ \left( \Deriv{\vec{v}}{t}\ - \vec{b} \right) \cdot \vec{v}^{\psi}} =
  \Integral{2}{\vec{t}\ \cdot \vec{v}^{\psi}} - \Integral{3}{\tens{\sigma} \cdot
    \tens{\varepsilon}^{\psi}},
\end{equation}\\
therefore (\ref{eq:BalanceMomentum_wf}) is the balance of momentum formulated in
its weak form.

@@

Finally, to solve the equation \eqref{eq:particle_balance_forces3}, a second order
temporal integration scheme is required. Therefore, time is
discretized in to a finite set of time steps $k = 1\ldots ,Nt$, where $k$ is the current time step and $N_t$
is the total number of time steps. Once the nodal equilibrium equation it is solved, the values in the nodes are
interpolated back in to the particles and each particle it is advected
to the new position,
\begin{equation}
  \label{eq:Updated_Lagrangian}
  \dot{\vec{v}}_p = \sum^{N_n}_{I}N_{Ip}\ \vec{a}_{I},\quad and\quad
  \dot{\vec{x}}_{p} = \sum^{N_n}_{I}N_{Ip}\ \vec{v}_{I}  
\end{equation}
In the MPM literature, this equations
(\ref{eq:particle_balance_forces3}) and (\ref{eq:Updated_Lagrangian}),
are solved with a forward Euler algorithm. In the next section, an
alternative time integration scheme based on the Newmark a-form is proposed.

@@

The time integration scheme of the MPM has been described in detail by many researchers
\cite{Sulsky1994}, \cite{Bardenhagen2002}, \cite{thesis_Andersen_2009} and summarized
in Figure \ref{fig:MPM_algorithm}. Other authors have proposed many
others time integration alternatives like
\cite{Guilkey2003}, \cite{Tran2019e}, \cite{Charlton2017}. In the first publication on
the MPM \cite{Sulsky1994}, the nodal acceleration in employed to
update the particles as
\begin{equation}
  \label{eq:Sulsky-1994-UL-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{a}_{I}^{k}
\end{equation}
\begin{equation}
  \label{eq:Sulsky-1994-UL-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + \Delta t\ \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{v}_{I}^{k}.
\end{equation}
However, as Andersen (2009)\cite{thesis_Andersen_2009} point out, this
algorithm has been shown to be numerically unstable due to that
$\vect{f}_I^{int,k}$ can be finite for an infinitesimal nodal mass
$\tens{m}$. This can lead to numerical issues when nodal acceleration
is obtained for evaluating \eqref{eq:Sulsky-1994-UL-x},\eqref{eq:Sulsky-1994-UL-v}. Hence, a
corrected version of this algorithm in shown in Zhang {\it et al.}
(2016)\cite{Zhang_book_2016}
\begin{equation}
  \label{eq:Zhang-2016-UL-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + \Delta t\ \sum^{N_n}_{I} \frac{N_{Ip}^{k}\ \vec{p}_{I}^{k}}{\tens{m}_I}.  
\end{equation}
\begin{equation}
  \label{eq:Zhang-2016-UL-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I}\frac{N_{Ip}^{k}\ \vec{f}_{I}^{k}}{\tens{m}_I},
\end{equation}
Later Tran \& Solowski (2019)\cite{Tran2019e} presented a
generalized-$\alpha$ scheme for the MPM inspired in the explicit time
integration algorithm proposed by Chung \& Hulbert
(1993)\cite{Geranlized_alpha_1993}, but with the particularity that
the acceleration is evaluated both in the beginning and the end of the
time step.
\begin{equation}
  \label{eq:Tran-2019-GA-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I} N_{Ip}^{k}\ \left[(1 - \gamma)\ \vect{a}_I^{k} +
    \gamma\ \vect{a}_I^{k+1} \right],\\
\end{equation}
\begin{equation}
\label{eq:Tran-2019-GA-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + 
  \sum^{N_n}_{I}N_{Ip}^{k} \left[ \Delta t\ \vec{v}_{I}^{k}+ \Delta t^2\left( (\frac{1}{2} - \beta)\
    \vec{a}_{I}^{k} + \beta\ \vec{a}_{I}^{k+1} \right) \right]
\end{equation}
\begin{equation}
  \label{eq:Tran-2019-GA-a}
  \vect{a}_p^{k+1} = \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{a}_{I}^{k+1}.
\end{equation}

@@

\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/Velocity-Comparative}
  }
  \caption{Velocity evolution at the point in the bar left side.}
  \label{fig:Dyka-MPM-uGIMP-LME}
\end{figure}

@@

For the derivation of this analytical solution we will consider  the
dynamic behaviour of a 1D elastic bar. The governing equations are the
following: (i) The balance of linear momentum,
\begin{equation}
  \label{eq:1D-balance-linear-momentum}
  \rho\ \Deriv{v}{t} = \Deriv{\sigma}{x} + \rho\ b,
\end{equation}
where $\sigma$ is the stress value, $\rho$ is the density,
$v$ is the velocity, and $b$ are the body
forces. (ii) The constitutive model, which for convenience of the
following developments will be written in terms of displacement and
velocities as, 
\begin{equation}
  \label{eq:1D-constitutive-equation}
  \Deriv{\sigma}{t} = E \Deriv{\varepsilon}{t},
\end{equation}
where $E$ is the elastic modulus. (iii) The compatibility equation
also in terms of the velocity field,
\begin{equation}
  \label{eq:CompatibilityEquation_e}
  \Deriv{\varepsilon}{t} = \Deriv{v}{x}.
\end{equation}
Next for simplicity, we remove the body forces from
\eqref{eq:1D-balance-linear-momentum}, and we will introduce
\eqref{eq:CompatibilityEquation_e} in
\eqref{eq:1D-constitutive-equation}, so we get the following system of equations,
\begin{align}
  \label{eq:1D-balance-linear-momentum-II}
  \Deriv{v}{t} &= \frac{1}{\rho}\ \Deriv{\sigma}{x}, \\
  \label{eq:1D-constitutive-equation-II}
  \Deriv{\sigma}{t} &= E\ \Deriv{v}{x}.
\end{align}

@@

The boundary conditions are in both cases of radiation as there is not
wave in-going from the exterior. So for the right side (fixed
boundary) we get the following conditions:
\begin{equation*}
  \Re^{II} = 0 \quad and \quad v_{x=L} = 0
\end{equation*}
Therefore $\sigma_{x=L} = -2*\sqrt{\rho E}\Re^{I}$. And in the left side
(free boundary) we get the following conditions:
\begin{equation*}
  \Re^{I} = 0 \quad and \quad \sigma_{x=0} = 0
\end{equation*}
Therefore $v_{x=0} = 2\Re^{II}$. Finally, applying this conditions in
the elastic bar sketched in \ref{fig:Dyka_Bar}, is possible to obtain
the velocity history in the right side of the bar
\ref{fig:vel_analytics_dyka} and the stress in the last quarter side of the Dyka
bar \ref{fig:stress_analytics_dyka} as is demanded in \cite{Dyka1995}.

>>>>

\title{Novel improvements in the Material Point Method to face
  dynamic problems: 
  Local-maximum entropy approximation and an explicit   predictor-corrector scheme. \thanks{Funding: This
    study was funded by Agustín de Betancourt Foundation (grant number
    262390106114).}
}
% \subtitle{}

\titlerunning{Novel improvements in the Material Point Method to face
  dynamic problems.} % if too long for running head

@@

\section{Introduction}
\label{intro}
Since the proposal of the Material Point Method (\textbf{MPM}) by
Sulsky {\it  et al.} (1994)\cite{Sulsky1994} as a generalization to
solids of the fluid implicit particle (FLIP) method
\cite{Brackbill1986}. It popularity has increased due to its ability
to deal with large strain regime without suffer mesh distortion
inaccuracies. However, this method suffers other kind of
instabilities, such those when material points crossing cell
boundaries. This give rise to the development of other interpolation
techniques to overcome this limitation such as the generalized
interpolation material point method (GIMP) Bardenhagen \& Kober
(2004)\cite{Bardenhagen2004}, the dual
domain material point method (DDMP) Zhang {\it et al.}
(2011)\cite{Zhang2011a}, the B-Spline MPM Tran {\it et al.}
(2019)\cite{Tran2019a} or the Conservative Taylor
Least Squares reconstruction Wobbes {\it et al.}
(2018)\cite{E_Wobbes_2018}. This techniques are devoted to mitigate the
``grid crossing'' error.
Nevertheless, in the presence of shock waves spurious
numerical noises appears despite of this using techniques \cite{Tran2019e}. These numerical
inaccuracies also known as wiggles are due to the time discretization
technique. A simple approach to face those spurious noises is to add a
damping source to the equilibrium equations. To overcome this
limitation, many researchers has proposed alternative time integration
schemes which reduce them by filtering high-frequencies. Some of these
attempts are the implicit GIMP (iGIMP) Charlton {\it et al.}
(2017)\nocite{Charlton2017} or the generalised-$\alpha$ scheme for the
MPM proposed by Tran \& Solowski (2019)\nocite{Tran2019e}.

The aim of this paper is to mitigate this spurious oscillations by the
employ of a suitable combination of the maximum-entropy (or local
\textit{max-ent}) $\gamma$ family shape functions, and the proposal of a explicit 
predictor-corrector scheme. Local \textit{max-ent} shape functions was
first introduced by Arroyo \& Ortiz (2006)\cite{Arroyo2006}, it belongs to the class of convex 
approximation schemes and provides a seamless transition between
finite elements (\textbf{FE}) and mesh-free interpolations. The
approximation scheme is based on a compromise between minimizing the
width of the shape function support and maximizing the information
entropy of the approximation. The local \textit{max-ent} approximation
may be regarded as a regularization, or \textit{thermalization}, of
Delaunay triangulation which effectively resolves the degenerate cases
resulting from the lack of uniqueness of the triangulation. Local
\textit{max-ent} basis functions possess many desirable properties for
mesh-free algorithms. First of all, they are entirely defined by the
nodal set and the domain of analysis. They are also non-negative,
satisfy the partition of unity property, and provide an exact
approximation for affine functions \cite{Arroyo2006}. Furthermore, as
it is not based in a voxel discretization as the GIMP, it is no prone to suffer
voxel domains overlap, which can introduce severe inaccuracies. This
approximation scheme has been proof to have a good performance under
the dynamic regime by other researchers like Navas {\it et al.}
(2018)\cite{Navas2018a} and Li {\it et al.} (2012)\cite{Li2012} for
Optimal Transportation Meshfree (\textbf{OTM}) method. And more
recently under the MPM framework by Wobbes {\it et
  al.}(2020)\cite{Wobbes2020} but limiting the $\gamma$ parameter to
zero. On the other hand, the time integration scheme here proposed is an accommodation of
the Newmark a-form ($\gamma = 0.5$ and $\beta = 0$)\footnote{Notice
that these $\gamma$ and $\beta$ parameters does not have the same
meaning for the LME shape functions.} which is the central difference
explicit (CD). We have choose this method among other suitable
alternatives as Wilson {\it et al.} (1972)\nocite{Wilson1972} or Chung
\& Hulbert (1993)\nocite{Geranlized_alpha_1993} because it simplicity
and it good performance dealing with solids dynamics problems like those
presented in \cite{Navas2018a}.

The article is organized as follows. Section \ref{sec:Notation}
briefly reviews the notation here employed. Next, Section \ref{sec:derivation-mpm}
is devoted to present briefly the governing equations of the elastic
problem, the variational formulation and the Galerkin procedure. In Section \ref{sec:epc-algor-mpm} an explicit predictor-corrector
time integration scheme for the MPM is proposed. Section
\ref{sec:local-max-ent} briefly reviews the local \textit{max-ent}
basis functions here employed. In Section \ref{sec:dynamic-benchmark}
a comparison between both methods MPM and OTM employing \textit{max-ent} shape
function is performed as well a parametric study for the $\gamma$
parameter. Both techniques will face to solve a classic dynamic
benchmark proposed by Dyka \& Ingel (1995)\cite{Dyka1995}. Finally,
conclusions and future research topics are exposed in Section \ref{sec:conclusions}.

@@

\section{Derivation of the MPM}
\label{sec:derivation-mpm}
The aim of this section is to provide an overview of
the standard explicit MPM algorithm \cite{Sulsky1994}. Therefore, it
is structured in the following shape: we will first give a
introduction to the governing equations \ref{sec:governing-equations},
later the variational statement of the problem will be presented \ref{sec:variational-formulation}, and
finally a discretization procedure thorough Galerkin is performed \ref{sec:Galerkin-procedure}.

The method has three main steps: (i) a variational recovery process,
where particle data is projected to the grid nodes, (ii) an Eulerian
step, where balance of momentum equation is expressed as a nodal
equilibrium equation thorough a FEM-like procedure, and finally (iii) a Lagrangian advection of the particles.

\subsection{Governing equations }
\label{sec:governing-equations}
In the MPM approach a continuum is considered. Let define a
$\varOmega$ domain occupied by an elastic body, and $\partial
\varOmega$ the boundary of the domain, where $\partial
\varOmega = \Gamma_d \bigcup \Gamma_n$ and $\Gamma_d \cap  \Gamma_n =
\emptyset$. Note that $\varGamma_d$ is reserved for Dirichlet boundary
conditions, and  $\varGamma_n$ for the Neumann. The behaviour of the continuum can be described by a set of governing equations which are
(i) the balance of momentum equation
\begin{equation}
  \label{eq:Balance-momentum}
\rho \vect{a} = \Div{\tens{\sigma}} + \rho \vect{b},
\end{equation}
(ii) the compatibility equation with ensure that the velocity
field is compatible with the Cauchy strain field,
\begin{equation}
  \label{eq:Compatibility-equation}
  \dot{\tens{\varepsilon}} = \GradS{\vect{v}} = \frac{1}{2}\left(
    \Grad{\vect{v}} +  \GradT{\vect{v}} \right),
\end{equation}
(iii) the constitutive equation, which relates the rate of Cauchy strain
tensor with the rate of Cauchy stress tensor,
\begin{equation}
  \label{eq:Constitutive-equation}
\dot{\tens{\sigma}} = \tens{D} \colon \dot{\tens{\varepsilon}},
\end{equation}
and finally (iv) the mass conservation, which can be obtained by
setting to zero the total derivative of the density field,
\begin{equation}
  \label{eq:Rho-material-derivative}
  \frac{D \rho}{D t} = \dot{\rho} + \rho \Div\vect{v}} = 0.
\end{equation}

\subsection{Variational formulation}
\label{sec:variational-formulation}
To write the variational statement of the problem, let us define a
virtual velocity field such that
\begin{equation}
  \label{eq:Hilbert-space}
  \vect{v}^{\psi} \in \mathcal{H}^1_0(\Omega) = \{ \vect{v}^{\psi} \in
  \mathcal{H}^1 \mid \vect{v}^{\psi} = \vect{0}\ \text{on}\ \Gamma_d \}.
\end{equation}
And which satisfies that the Cauchy sequences are convergent in $\varOmega$
\begin{equation}
  \label{eq:cauchy-secuence}
  \Integral{3}{\vect{v}^{\psi}} < \infty\ \quad\text{and}\quad \Integral{3}{\tens{\varepsilon}^{\psi}} < \infty
\end{equation}
The principle of virtual work states that the equilibrium solution to
the boundary value problem of elasticity is the function $\vect{v} \in
\mathcal{H}^1_0$ such that, for $\vect{v}^{\psi} \in
\mathcal{H}^1_0$, the following holds:
\begin{equation}
  \label{eq:BalanceMomentum_wf}
  \Integral{3}{\rho\ \left( \Deriv{\vec{v}}{t}\ - \vec{b} \right) \cdot \vec{v}^{\psi}} =
  \Integral{2}{\vec{t}\ \cdot \vec{v}^{\psi}} - \Integral{3}{\tens{\sigma} \cdot
    \tens{\varepsilon}^{\psi}},
\end{equation}\\
therefore (\ref{eq:BalanceMomentum_wf}) is the balance of momentum formulated in
its weak form.

@@

Finally, to solve the equation \eqref{eq:particle_balance_forces3}, a second order
temporal integration scheme is required. Therefore, time is
discretized in to a finite set of time steps $k = 1\ldots ,Nt$, where $k$ is the current time step and $N_t$
is the total number of time steps. Once the nodal equilibrium equation it is solved, the values in the nodes are
interpolated back in to the particles and each particle it is advected
to the new position,
\begin{equation}
  \label{eq:Updated_Lagrangian}
  \dot{\vec{v}}_p = \sum^{N_n}_{I}N_{Ip}\ \vec{a}_{I},\quad and\quad
  \dot{\vec{x}}_{p} = \sum^{N_n}_{I}N_{Ip}\ \vec{v}_{I}  
\end{equation}
In the MPM literature, this equations
(\ref{eq:particle_balance_forces3}) and (\ref{eq:Updated_Lagrangian}),
are solved with an explicit forward Euler algorithm. In the next
section, an alternative time integration scheme based on the Newmark a-form is presented.

@@

The time integration scheme of the MPM has been described in detail by many researchers
\cite{Sulsky1994}, \cite{Bardenhagen2002}, \cite{thesis_Andersen_2009} and summarized
in Figure \ref{fig:MPM_algorithm}. Other authors have proposed many
others time integration alternatives like
\cite{Guilkey2003}, \cite{Tran2019e}, \cite{Charlton2017}. In the first publication on
the MPM \cite{Sulsky1994}, the nodal acceleration was employed to
update the particles as
\begin{equation}
  \label{eq:Sulsky-1994-UL-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{a}_{I}^{k}
\end{equation}
\begin{equation}
  \label{eq:Sulsky-1994-UL-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + \Delta t\ \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{v}_{I}^{k}.
\end{equation}
However, as Andersen (2009)\cite{thesis_Andersen_2009} point out, this algorithm has been shown to be numerically unstable due to that
$\vect{f}_I^{int,k}$ can be finite for an infinitesimal nodal mass
$\tens{m}$. This can lead to numerical issues when nodal acceleration
is obtained for evaluating \eqref{eq:Sulsky-1994-UL-x},\eqref{eq:Sulsky-1994-UL-v}. Hence, a
corrected version of this algorithm in shown in Zhang {\it et al.}
(2016)\cite{Zhang_book_2016}
\begin{equation}
  \label{eq:Zhang-2016-UL-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + \Delta t\ \sum^{N_n}_{I} \frac{N_{Ip}^{k}\ \vec{p}_{I}^{k}}{\tens{m}_I}.  
\end{equation}
\begin{equation}
  \label{eq:Zhang-2016-UL-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I}\frac{N_{Ip}^{k}\ \vec{f}_{I}^{k}}{\tens{m}_I},
\end{equation}
Later Tran \& Solowski (2019)\cite{Tran2019e} presented a
generalized-$\alpha$ scheme for the MPM inspired in the explicit time
integration algorithm proposed by Chung \& Hulbert
(1993)\cite{Geranlized_alpha_1993}, but with the particularity that
the acceleration is evaluated both in the beginning and the end of the
time step.
\begin{equation}
  \label{eq:Tran-2019-GA-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I} N_{Ip}^{k}\ \left[(1 - \gamma)\ \vect{a}_I^{k} +
    \gamma\ \vect{a}_I^{k+1} \right],\\
\end{equation}
\begin{equation}
\label{eq:Tran-2019-GA-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + 
  \sum^{N_n}_{I}N_{Ip}^{k} \left[ \Delta t\ \vec{v}_{I}^{k}+ \Delta t^2\left( (\frac{1}{2} - \beta)\
    \vec{a}_{I}^{k} + \beta\ \vec{a}_{I}^{k+1} \right) \right]
\end{equation}
\begin{equation}
  \label{eq:Tran-2019-GA-a}
  \vect{a}_p^{k+1} = \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{a}_{I}^{k+1}.
\end{equation}

@@

\begin{figure*}\sidecaption
  \centering
  \resizebox{0.7\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/Velocity-Comparative}
  }
  \caption{Velocity evolution at the point in the bar left side.}
  \label{fig:Dyka-MPM-uGIMP-LME}
\end{figure*}

@@

For the derivation of this analytical solution we will consider  the
dynamic behaviour of a 1D elastic bar. The governing equations are the
following: (i) The balance of linear momentum,
\begin{equation}
  \label{eq:1D-balance-linear-momentum}
  \rho\ \Deriv{v}{t} = \Deriv{\sigma}{x},
\end{equation}
where $\sigma$ is the stress value, $\rho$ is the density, and 
$v$ is the velocity. (ii) The constitutive model, which for convenience of the
following developments will be written in terms of displacement and
velocities as, 
\begin{equation}
  \label{eq:1D-constitutive-equation}
  \Deriv{\sigma}{t} = E \Deriv{\varepsilon}{t},
\end{equation}
where $E$ is the elastic modulus. (iii) The compatibility equation
also in terms of the velocity field,
\begin{equation}
  \label{eq:CompatibilityEquation_e}
  \Deriv{\varepsilon}{t} = \Deriv{v}{x}.
\end{equation}
Next for simplicity, we will introduce \eqref{eq:CompatibilityEquation_e} in 
\eqref{eq:1D-constitutive-equation}, so we get the following system of equations,
\begin{align}
  \label{eq:1D-balance-linear-momentum-II}
  \Deriv{v}{t} &= \frac{1}{\rho}\ \Deriv{\sigma}{x}, \\
  \label{eq:1D-constitutive-equation-II}
  \Deriv{\sigma}{t} &= E\ \Deriv{v}{x}.
\end{align}

@@

The boundary conditions are in both cases of radiation as there is not
wave in-going from the exterior. So for the right side (fixed
boundary) we get the following conditions:
\begin{equation*}
  \Re^{II} = 0 \quad and \quad v_{x=L} = 0
\end{equation*}
Therefore $\sigma_{x=L} = -2\sqrt{\rho E}\ \Re^{I}$. And in the left side
(free boundary) we get the following conditions:
\begin{equation*}
  \Re^{I} = 0 \quad and \quad \sigma_{x=0} = 0
\end{equation*}
Therefore $v_{x=0} = 2\Re^{II}$. Finally, applying this conditions in
the elastic bar sketched in \ref{fig:Dyka_Bar}, is possible to obtain
the velocity history in the right side of the bar
\ref{fig:vel_analytics_dyka} and the stress in the last quarter side of the Dyka
bar \ref{fig:stress_analytics_dyka} as is demanded in \cite{Dyka1995}.
